\tit

### 2021-02-20 paper.tex
predicate operations spacing
proofreading pass

%% I prefer ∅ in XITSMath-Regular, but I don't know how to make setmathfont
%% work with acmart.
\ifacm
\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
\newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
\else
\setmathfont{XITSMath-Regular.otf}
\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
\newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
\fi

@@

Languages are usually formalized either set-theoretically as sets of strings or operationally as parsers.
Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters'').
\figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).

The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and elements/values of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2008AFP, BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
Specifically,
\begin{itemize}

@@

\item For a function {\AB F \AS : \AB A \AS → \APT{Set}}, the types {\AF ∃ \AB F}  and {\AF ∀ \AB x \AS → \AB F \AB x} represent existential and universal quantification as \emph{dependent} product and function types.

@@

Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsing}.
(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{P}}.

The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
As described in \secref{Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis, with compelling and correct implementations following easily and systematically as corollaries.

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated (and proved) about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions.
\item
  A duality is revealed between regular expressions and tries in the context of language differentiation.
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
\item
  The vocabulary and reasoning of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary.
  This generalization consists of two generally useful covariant applicative functors on functions (the usual one on functions-from-X and another on functions-to-Set).
  Properties are proved in terms of this more general vocabulary.
\item
  Properties are defined first at the level of predicates, making both easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported effortlessly to the computable parsing implementations.
\end{itemize}
The source code for this paper is freely available and contains fully verified Agda code.
All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.

@@

In order to convert languages into parsers, it will help to decompose languages into a regular form.
Since lists are defined by an algebraic data type, we can decompose languages---and indeed any function from lists---according to constructor, as follows:
\agda{νδ}
In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}, i.e., the tails of strings in \AB{P} that start with \AB{a}.

@@

As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler.
The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.

@@

For an effective implementation, we will have to bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}).

@@

\item Given two functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
In this use, {\AB C} is \AD ⊥, since we are proving a negation.

\item The \AF{\_✶‽} definition reflects the fact that the empty list exists for every element type (even \AF ⊥).

@@

The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
An automatic solution would need some form of \emph{reflection} of language operations or \AF{ν} and \AF{δ} as inspectable data.

This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.

Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{The classic trie (``prefix tree'') data structure was introduced by Thue in 1912 to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary (regular) algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.

Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), and this time the compiler successfully proves \emph{total} correctness (including termination).

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids, while \AF{`} is specific to lists.

\figrefdef{predicates}{Predicate operations}{
\ifacm
\small
\fi
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{-2ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{17em}
%% Transform codomain: %\small
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{20em}
%% Transform domain: %\small
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set⇂} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} the type \AF{Set⇂}.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{-2ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which also includes language complement (\AF ∁).

\rnc\source{Calculus}
The \AF{ν} and \AF{δ} lemmas for codomain operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
(In contrast, the \AF{ν} and \AF{δ} lemmas do not easily generalize for domain operations.)
Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation with broader coverage code.

@@

Conveniently, booleans and types form commutative semirings with all necessary definitions already in the standard library.
(The equivalence relation used for types is isomorphism rather than equality.)
Both are also closed.
For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.

This first closed semiring for predicates follows from a much more general pattern.
Given any two types \AB{A} and \AB{B}, if \AB{B} is a monoid then {\AB A \AS → \AB B} is as well.
The monoid operation \AB{\_∙\_} lifts to the function-level binary operation {\AS λ (\AB{f} \AB{g} : \AB{A} \AS → \AB{B}) \AS → \AS λ \AB a \AS → \AB f \AB a \AB ∙ \AB g \AB a}.
The monoid identity {\AB ε \AS : \AB{B}} lifts to the identity {\AS λ \AB a \AS → \AB ε}.
All of the laws transfer from \AB{B} to {\AB A \AS → \AB B}.
Likewise for other algebraic structures.

Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses.
Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.

>>>>

%% I prefer ∅ in XITSMath-Regular, but I don't know how to make setmathfont
%% work with acmart.
\ifacm
\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
\newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
\else
\setmathfont{XITSMath-Regular.otf}
\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
\newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
\fi

@@

Languages are usually formalized either set-theoretically as sets of strings or operationally as parsers.
Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters'').
The usual language operations are defined in \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}}, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).

The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and element (values) of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2008AFP, BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
Specifically,
\begin{itemize}

@@

\item For a function {\AB F \AS : \AB A \AS → \APT{Set} \AB{ℓ}}, the types {\AF ∃ \AB F}  and {\AF ∀ \AB x \AS → \AB F \AB x} represent existential and universal quantification as \emph{dependent} product and function types.

@@

Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{P}}.

The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
As described in \secref{Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis for specification and reasoning, with correct and computable parsing implementations following systematically as corollaries.

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
\item
  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secreftwo{Symbolic Differentiation}{Automatic Differentiation}).
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
\item
  The vocabulary of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
  % Properties are proved in terms of this more general vocabulary.
\item
  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported effortlessly to the computable parsing implementations (\secref{Transporting Properties}).
\end{itemize}
Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.
All type signatures and definitions in the paper are generated by the Agda compiler.

@@

In order to convert languages into parsers, it will help to decompose languages into a regular form.
Since lists are defined by an algebraic data type, we can decompose languages---and indeed any function from lists---according to constructor, as follows:
\agda{νδ}
In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}\out{, i.e., the tails of strings in \AB{P} that start with \AB{a}}.

@@

The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler.

@@

For an effective implementation, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}),  sometimes called ``\AF{absurd}'' or ``\AF{⊥-elim}''.

@@

\item For functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
% In this use, {\AB C} is \AD ⊥, since we are proving a negation.

\item \AF{\_✶‽} reflects the fact that the empty list exists for every element type (even \AF ⊥).

@@

The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
An automatic solution would need some form of \emph{reflection} of language operations or of \AF{ν} and \AF{δ} as inspectable data.

This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
Fortunately, there are two standard solutions for \emph{computable} differentiation, commonly referred to as ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those symbolic representations.
The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
This latter option is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a function {\AF{⟦\_⟧}} that converts a symbolic language representation to a computable parser.

Decidable parsing (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would.)

In \figref{symbolic-defs}, note the role of the isomorphisms from \figref{nu-delta-lemmas}.
Since those isomorphisms relate \AF{ν} and \AF{δ} on the language-building operations to types and languages expressed in that \emph{same} vocabulary, we end up with a recursive algorithm.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

The correctness proof is \emph{almost} accomplished by type- and termination-checking, but a technical problem arises.
The {\AF{δ} (\AB{p} \AF ⋆ \AB{q})} clause does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
(\figref{automatic-defs} compiles only due to suppressing termination checking for the problematic clause with a compiler pragma.)
Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
\rnc\source{SizedAutomatic}
This solution only requires giving \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.
Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), with the sole exception of removing the compiler pragma that suppressed termination checking.
This time the compiler successfully proves \emph{total} correctness (including termination).

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids\footnote{The \emph{definitions} of those operations do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).}, while \AF{`} is specific to lists.

\figrefdef{predicates}{Predicate operations}{
\ifacm
\small
\fi
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{-2ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set⇂} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} the type \AF{Set⇂}.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{-2ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).

\rnc\source{Calculus}
The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.

@@

Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.
(The equivalence relation used for types is isomorphism rather than equality.)
Both are also closed.
For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.

This first closed semiring for predicates follows from a much more general pattern.
Given any two types \AB{A} and \AB{B}, if \AB{B} is a monoid then {\AB A \AS → \AB B} is as well.
The monoid operation \AB{\_∙\_} lifts to the function-level binary operation {\AS λ (\AB{f} \AB{g} : \AB{A} \AS → \AB{B}) \AS → \AS λ \AB a \AS → \AB f \AB a \AB ∙ \AB g \AB a}.
The monoid identity {\AB ε \AS : \AB{B}} lifts to the identity {\AS λ \AB a \AS → \AB ε}.
All of the laws transfer from \AB{B} to {\AB A \AS → \AB B}.
Likewise for other algebraic structures.
(As always, proofs are in this paper's source code.)

Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses.
Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling defined by conjoining membership proofs (as in \figref{lang-via-pred}).
